Lenses with a Difference

### 2018-06-15 papers/DifferenceLenses/main.tex
Made some changes

% % Abstract % Note: \begin{abstract}...\end{abstract} environment must come %
% before \maketitle command
\begin{abstract}
Many of the bidirectional (bx) programming languages that exist today focus on
restoring consistency between source and target data. Consequently, these languages often provide two guarantees: (1) that a transformation restores consistency when an
update is made to source or target data, and (2) that trivial updates to a the source result to trivial updates to the target and vice versa. Unfortunately, in situations
where there are many different ways of restoring consistency, these languages
do not constrain how a transformation chooses to do so. As a result, the transformations derived using these languages can lead to unexpected results. In this paper we present a formalization of bx transformations that enables a programmer to express non-trivial guarantees on update propagation. We will then use this formalization of bx transformations to describe a compositional language of {\em difference lenses}: each difference lens gives non-trivial guarantees on how it propagates non-trivial updates. We will then embed various lens programming languages into the language of difference lenses, and describe non-trivial guarantees that these languages provide beyond consistency restoration.
\end{abstract}

@@

For example, assume that a source consists of a list of pairs of names, and that the target consists of a list of single names. Assume that $\ell$ is a bx which in the forward direction (i.e. the \get direction) simply discards the first of the two names:
$$\ell.\get \; [``Ada \; Lovelace", \; ``Alan \; Turing"] = [``Lovelace", \; ``Turing"]$$
If the order of the names in the target is reversed then a user might expect that in the backward direction (i.e. the \pput direction), the lens would simply restore the original names:
$$\ell.\pput \; [``Ada \; Lovelace", \; ``Alan \; Turing"] \; [``Turing", \; ``Lovelace"] = [``Ada \; Lovelace", \; ``Alan \; Turing"]$$
However, $\ell$ may not perform this alignment so that instead, the following result occurs:
$$\ell.\pput \; [``Ada \; Lovelace", \; ``Alan \; Turing"] \; [``Turing", \; ``Lovelace"] = [``Ada \; Turing", \; ``Alan \; Lovelace"]$$
Consider another example where a source consists of a list of names and a number while the target consists of a list of names. Assume that $\ell$ is a bx which in the \get direction simply discards the number:
$$\ell.\get \; ([``Ada \; Lovelace", \; ``Alan \; Turing"], \; 2)  = [``Lovelace", \; ``Turing"]$$
A user might expect that $\ell$ should always restore the number in the \pput direction so that the following result occurs:
$$\ell.\pput \; ([``Ada \; Lovelace", \; ``Alan \; Turing"], \; 2) \; [``Lovelace"] = ([``Ada \; Lovelace"], \; 2)$$
However, perhaps the intended semantics of the source is that the number which is the second component of a source denotes the length of the list in the first component of the source so that an application of \pput leads to the following result:
$$\ell.\pput \; ([``Ada \; Lovelace", \; ``Alan \; Turing"], \; 2) \; [``Lovelace"] = ([``Ada \; Lovelace"], \; 1)$$
In both of these examples, there are multiple ways to restore consistency; that is, there are multiple ways to define $\ell.\pput \; s \; v$ in such a way that $\ell.\get \; (\ell.\pput \; s \; v) = v$. Moreover, the source that results from a \pput application may be vastly different from what a user expects, or may break an invariant that is not explicitly maintained by the source.

@@

We take the approach that a bx should not have to explicitly propagate updates when going from the source to the target and vice versa. Instead, the programmer should be able provide guarantees on how a bx propagates updates. For instance, in the second example, let $\delta_S$ be a function which takes two source elements $(\ell_1, n_1)$ and $(\ell_2, n_2)$ and returns the number $n_1 - n_2$, and let $\delta_T$ be a function which takes two target lists $\ell_1$ and $\ell_2$ and returns the number $length(\ell_1) - \length(\ell_2)$. Then the programmer should be able to express the fact that the bx restores the length of the list in the backward direction, which is expressed using the following equation:
$$\delta_S((\ell_1, n_1), \ell.\pput \; (\ell_1, n_1) \; \ell_2) = \delta_V(\ell.\get \; (\ell_1, n_1), \ell_2) $$
In this paper, we introduce a formalization of bx transformations that enables a programmer to express guarantees such as the one above and show how our formalism can be used to characterize update propagation behaviour of bidirectional transformations.

The major technical contributions that we will make in this paper are the following:
\begin{enumerate}
  \item
  We present a formalization of bx transformations that enables a programmer to express non-trivial update propagation behaviour.
  \item
  We use our formalization to develop a compositional language of {\em difference lenses} based on the {\em basic lenses} implemented in the lens programming language Boomerang \cite{foster2009bidirectional}, and describe non-trivial update propagation guarantees that these lenses provide.
  \item
  We show how to represent matching lenses and  relational lenses \cite{bohannon2006relational} as difference lenses, and hence describe non-trivial guarantees that these lens languages provide as well.
\end{enumerate}
\iffalse
 we describe a compositional language of lenses called {\em difference lenses}, where each difference $\ell$ satisfies the following equation,
\begin{equation}\label{difflenslaw}
\delta_S(s, \ell.\pput \; s \; v) = \delta_V(\ell.\get \; s, v)
\end{equation}
for appropriate real-valued $\delta_S : S \times S \longrightarrow \mathbb{R}^{\geq 0}$ and $\delta_S : S \times S \longrightarrow \mathbb{R}^{\geq 0}$.

>>>>

% % Abstract % Note: \begin{abstract}...\end{abstract} environment must come %
% before \maketitle command
\begin{abstract}
Many of the bidirectional (bx) programming languages that exist today focus on
restoring consistency between source and target data. Consequently, these languages often provide two guarantees: (1) that a transformation restores consistency when an
update is made to source or target data, and (2) that trivial updates to a the source result to trivial updates to the target and vice versa. Unfortunately, in situations
where there are many different ways of restoring consistency, these languages
do not constrain how a transformation chooses to do so. As a result, the transformations derived using these languages can lead to unexpected results. In this paper we present a formalization for bx transformations that requires a bx to provide guarantees on how it propagates updates. We will then describe a specific bx language which we call {\em difference lenses}; this language is based on the {\em basic lenses} defined in the bx programming language {\em Boomerang}, but each difference lens comes with explicit guarantees on update propagation. We will then express various lens programming languages as difference lenses, and hence describe non-trivial guarantees on update propagation that these languages provide as well.
\end{abstract}

@@

For example, assume that a source consists of a list of pairs of names, and that the target consists of a list of single names. Assume that $\ell$ is a bx which in the forward direction (i.e. the \get direction) simply discards the first of the two names:
$$\ell.\get \; [``Ada \; Lovelace", \; ``Alan \; Turing"] = [``Lovelace", \; ``Turing"]$$
If the order of the names in the target is reversed then a user might expect that in the backward direction (i.e. the \pput direction), the lens would simply restore the original names:
$$\ell.\pput \; [``Ada \; Lovelace", \; ``Alan \; Turing"] \; [``Turing", \; ``Lovelace"] = [``Ada \; Lovelace", \; ``Alan \; Turing"]$$
However, $\ell$ may not perform this alignment so that instead, the following result occurs:
$$\ell.\pput \; [``Ada \; Lovelace", \; ``Alan \; Turing"] \; [``Turing", \; ``Lovelace"] = [``Ada \; Turing", \; ``Alan \; Lovelace"]$$
Consider another example where both a source and a target consists of a list where every element on the list is a name and a year. Assume that $\ell$ is a bx which in the \get direction returns all elements of the source list that have year greater than or equal to 1900:
$$\ell.\get \; [``Ada \; Lovelace, \; 1815", \; ``Alan \; Turing, \; 1912"] = [``Alan \; Turing, \; 1912"]$$
Suppose that the entry $``Ada \; Lovelace, \; 1915"$ is then added to the target list. Should $\ell.\pput$ replace the old entry containing the name $``Ada \; Lovelace"$,
\begin{align*}
&\ell.\pput \; [``Ada \; Lovelace, \; 1815", \; ``Alan \; Turing, \; 1912"] \; [``Alan \; Turing, \; 1912", \; ``Ada \; Lovelace, \; 1915"] \\
&= [``Ada \; Lovelace, \; 1915", \; ``Alan \; Turing, \; 1912"]
\end{align*}
or should the \pput application retain both versions of $``Ada \; Lovelace"$?
\begin{align*}
&\ell.\pput \; [``Ada \; Lovelace, \; 1815", \; ``Alan \; Turing, \; 1912"] \; [``Alan \; Turing, \; 1912", \; ``Ada \; Lovelace, \; 1915"] \\
&= [``Ada \; Lovelace, \; 1815", \; ``Alan \; Turing, \; 1912", \; ``Ada \; Lovelace, \; 1915"]
\end{align*}
In both of these examples, there are multiple ways to restore consistency---that is, there are multiple ways to define $\ell.\pput \; s \; v$ in such a way that $\ell.\get \; (\ell.\pput \; s \; v) = v$---and the results from the different \pput applications may be vastly different from what a user expects.

@@

We take the approach that a bidirectional programming language should have explicit guarantees on how programs written in that language propagate updates; ideally these guarantees should be strictly stronger than hippocraticness. This way a programmer who derives bx transformations in this language has a clearer understanding of the behaviour of the programs that she writes.

In this paper we will show how this can be done, by describing a formalization of bx transformations that explicitly accounts for update propagation, and presenting a series of examples showing how this principle can be applied to existing bx programming languages. More specifically, the major technical contributions that we will make in this paper are the following:
\begin{enumerate}
\item
We present a formalization for bx transformations that requires a bx to provide guarantees on how it propagates updates. 
\item
We describe a specific bx language which we call {\em difference lenses}; this language is based on the {\em basic lenses} defined in the bx programming language {\em Boomerang}\cite{foster2009bidirectional}, but each difference lens comes with explicit guarantees on update propagation. As a result of this work, we will describe non-trivial update propagation properties that Boomerang's basic lenses provide.
  \item
  We show how to represent matching lenses and  relational lenses \cite{bohannon2006relational} as difference lenses, and hence describe non-trivial guarantees that these lens languages provide as well.
\end{enumerate}
\iffalse
 we describe a compositional language of lenses called {\em difference lenses}, where each difference $\ell$ satisfies the following equation,
\begin{equation}\label{difflenslaw}
\delta_S(s, \ell.\pput \; s \; v) = \delta_V(\ell.\get \; s, v)
\end{equation}
for appropriate real-valued $\delta_S : S \times S \longrightarrow \mathbb{R}^{\geq 0}$ and $\delta_S : S \times S \longrightarrow \mathbb{R}^{\geq 0}$.
